A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-22 paper.tex
New local trees section

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
[NOTE: we could move this to the later section currently
called "the equivalence of ARGS and the local trees",and
refocus a bit to "ARGs and local trees".]
The relationship between an ARG and the corresponding local trees
is subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of a marginal distributions indicating
an integrated tree over many positions.

Firstly, the internal nodes in the local trees must be labelled
(i.e.\ marked with the identity of the corresponding ARG node).
If the nodes are not labelled, we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.
% JK: Does this need to be said?
% (One may use the time of a node as a proxy for identity,
% but it is not reliable as we can have multiple nodes at the
% same time, e.g. with time discretisation.)

The precision of the information about recombination that
can be represented in the trees depends the ARG itself.
Secondly, the local trees must contain ``unary'' nodes.
[Some discussion].
With these two conditions met, there is a direct correspondence
between a gARG and the trees.

>>>>

@@

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are said to be
``embedded'' within in ARG [CITES], and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of a marginal distributions indicating
an integrated tree over many positions.

It is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its parent ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree
and along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, there
is a one-to-one correspondance between an ARG and
the sequence of local trees that it encodes.

The internal nodes of trees, however, are usually
seen as \emph{unlabelled}, since they most often correspond
to anonymous hypothetical ancestors. Internal
ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.

@@
